---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r}
options(digits=4)
```

## Loading and preprocessing the data
The first step is to load and explore the data set. The data set is contained in a single CSV file, "activity.csv", delivered as a zipped file "activity.zip", containing a single file, "activity.csv".

It is instructive to then examine the structure of the dataset, and the first and last few rows.

```{r}
# Read raw data
unzip("./activity.zip")
activity <- read.csv("./activity.csv", colClasses=c("integer", "POSIXct", "integer"))

# Structure of the raw data
str(activity)

# First and last 6 records
head(activity)
tail(activity)
```

As shown in the output above, the data set consists of 17568 observations of only 3 variables:

- steps:    Number of steps taken in a given 5-minute interval.
- date:     Date at which the measurement was taken in yyyy-MM-dd format.
- interval: which 5 minute interval the measurement was taken in.

One can see that there are a significant number of observations with missing values for the "steps" variable.

## What is mean total number of steps taken per day?


```{r}
suppressMessages(library("dplyr"))

# Calculate Total Steps per day
steps_per_day <- activity %>% 
                    group_by(date) %>%
                    summarize(steps_per_day=sum(steps, na.rm=T)) %>%
                    select(date, steps_per_day)

# Calculate Mean and Median
spd_mean <- mean(steps_per_day$steps_per_day)
spd_median <- median(steps_per_day$steps_per_day)
```
Below is a histogram of the number of steps per day.

```{r}
# Histogram of steps per day
hist(steps_per_day$steps_per_day, xlab="Steps per day", main="Total number of steps per day.")
abline(v=spd_mean, lty=2, col="blue")
abline(v=spd_median, lty=2, col="red")
legend("topright", legend=c("mean", "median"), lty=2, col=c("blue", "red"))
```

The average number of steps taken per day is `r spd_mean`, while the median is `r spd_median`. 

## What is the average daily activity pattern?

```{r}

average_by_5min <- activity %>%
                    group_by(interval) %>%
                    summarize(avg_steps=mean(steps, na.rm=T))

plot(average_by_5min$avg_steps ~ average_by_5min$interval, type="l", xlab="Interval", ylab="Average Steps per Interval", main="Average steps per 5-min interval over all days.")
```

```{r}
top_5min_interval <- average_by_5min[average_by_5min$avg_steps==max(average_by_5min$avg_steps, na.rm=T),]$interval

top_5min_interval
```

The interval with the highest average number of steps across all days is `r top_5min_interval`

## Imputing missing values

The first step is to calculate the total number of observations with missing values for the "steps" variable:

```{r}

count.na <- count(activity[is.na(activity$steps),])
count.na
```

Total number of observations with missing step counts: `r count.na`, or `r count.na / count(activity) * 100`% of all observations.

Missing values are next filled in for a given interval on a given day by using instead the median value for that interval across all days.

```{r}
suppressMessages(library(dplyr))

# calculate median steps across all days for all intervals
median_by_5min <- activity %>%
                    group_by(interval) %>%
                    summarize(mdn_steps=median(steps, na.rm=T))

# merge the median into the main dataset.
activity_with_intervalmdn <- merge(activity, median_by_5min, by="interval", all.x=T, all.y=F)

# replace NAs with median
activity_with_intervalmdn[is.na(activity_with_intervalmdn$steps), ]$steps <-
  activity_with_intervalmdn[is.na(activity_with_intervalmdn$steps), ]$mdn_steps

# remove unwanted columns.
activity_imputed <- activity_with_intervalmdn %>% 
                      select(date, interval, steps)
```

Here is the structure of the new dataset, with missing values impured as discussed above, as well as its first and last few records:

```{r}
str(activity_imputed)
head(activity_imputed)
tail(activity_imputed)
```

Next we recalculate and display the total, median and mean number of steps per day, but this time using the dataset with missing values imputed.

```{r}
# Calculate the total number of steps per day using the imputed values.
steps_per_day_with_imputation <- activity_imputed %>% 
                    group_by(date) %>%
                    summarize(steps_per_day=sum(steps, na.rm=T)) %>%
                    select(date, steps_per_day)

# Calculate Mean and Median of steps per day with imputation.
spd_mean_with_imputation <- mean(steps_per_day_with_imputation$steps_per_day)
spd_median_with_imputation <- median(steps_per_day_with_imputation$steps_per_day)
```

The histogram as above, but after missing values are imputed.

```{r}
# Histogram of steps per day
hist(steps_per_day_with_imputation$steps_per_day, xlab="Steps per day", main="Total number of steps per day after Imputation.")
abline(v=spd_mean_with_imputation, lty=2, col="blue")
abline(v=spd_median_with_imputation, lty=2, col="red")
legend("topright", legend=c("mean", "median"), lty=2, col=c("blue", "red"))
```


The average number of steps per day after imputation of missing values is `r spd_mean_with_imputation`, having moved closer to the median (it was `r spd_mean` before missing values were filled in). The median is still `r spd_median_with_imputation`. This is to be expected since we used the median value to fill in blanks.

## Are there differences in activity patterns between weekdays and weekends?

The first step is to add a column identifying whether the date for an observation corresponds to a weekday or a weekend day. The structure of the new dataset follows below:

```{r}
activity_imputed_with_wkday <- activity_imputed %>%
  mutate(is.weekday=factor(ifelse(weekdays(date, abbreviate=T) %in% c("Sat", "Sun"), "weekend", "weekday")))

str(activity_imputed_with_wkday)
```

Next the average steps per interval is calculated, grouped by whether the value corresponds to a weekday or a weekend, in order to display a panel plot by day type of the relationship between interval and average number of steps.
```{r}
wkday_averages <- activity_imputed_with_wkday %>%
                    group_by(interval, is.weekday) %>%
                    summarize(avg_steps=mean(steps)) %>%
                    ungroup()


suppressWarnings(library(lattice))

with(wkday_averages, 
     xyplot(avg_steps ~ interval | is.weekday, panel=panel.lines, layout=c(1, 2), ylab="Number of steps", xlab="Interval"))
```